Staged Abstract Interpreters

### 2019-09-09 paper/40_absinterp.tex
polish

The co-inductive caching algorithm is implemented as an instrumentation over the
@eval@ function (Figure \ref{fig:coind_cache}), and it also closes the open recursion.
The instrumentation works as follows. Initially, the two caches are both empty.
During the iteration, we first check whether the @out@ cache contains the
configuration @cfg@, which represents the current desired computation. If @out@ does,
then the result is directly returned throughout the monad stack.
Otherwise, we first retrieve the result from @in@ (@⊥@ if @in@ not contains
@cfg@), and update this result from @in@ into the @out@ cache in the fashion of
join. Then, we invoke @ev@ to evaluate the result for this iteration, where
@ev@ takes @fix(ev)@ as the self reference.  After the evaluation, the store @σ@
may have been changed, so we obtain the latest store and construct the
result @(v, σ)@, which will be used to update the @out@ cache.

>>>>

The co-inductive caching algorithm is implemented as an instrumentation over the
@eval@ function (Figure \ref{fig:coind_cache}), and it also closes the open recursion.
The instrumentation works as follows. Initially, the two caches are both empty.
During the iteration, we first check whether the @out@ cache contains the
configuration @cfg@, which represents the current desired computation. If @out@
does contain @cfg@, then the result is directly returned throughout the monad
stack.
Otherwise, we first retrieve the result from @in@ (@⊥@ if @in@ does not contain
@cfg@), and update this result from @in@ into the @out@ cache in the fashion of
join. Then, we invoke @ev@ to evaluate the result for this iteration, where
@ev@ takes @fix(ev)@ as the self reference.  After the evaluation, the store @σ@
may have been changed, so we obtain the latest store and construct the
result @(v, σ)@, which will be used to update the @out@ cache.

### 2019-09-09 paper/50_stageabs.tex
polish

The @ap_clo@ method for function applications is also similarly mixing the two
previous interpreters.  We use the staged version of @lift_nd@ to lift the
next-stage set of functions into the monad stack.
For each closure @clo@ in the set, we a generate next-stage value representing
the function application @clo(arg)@ by using @emit_ap_clo@.  Again, the method
@emit_ap_clo@ produces a current-stage representation of the future-stage
application result.  Finally, we reify the @out@ cache, store, and values,
which are all from the future-stage, back into the current-stage monadic value.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[ValSt], Cache)]
  def ap_clo(ev: Expr => Ans)(funs: Rep[Value], arg: Rep[Value]): Ans = for {
    σ   <- get_store;      clo <- lift_nd[AbsValue](funs)
    in  <- ask_in_cache;   out <- get_out_cache
    res <- lift_nd[(Set[ValSt], Cache)](Set(emit_ap_clo(clo, arg, σ, in, out)))
    _   <- put_out_cache(res._2);  vs <- lift_nd[ValSt](res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

@@

Particularly, if we are specializing a monovariant analysis, the address space
is equivalent to the set of all identifiers in the program. Utilizing this fact, the
result of accesses to the environment can be computed statically and we may
generate addresses directly during staging. After which, the store can be
specialized as arrays by using the way mentioned above.

>>>>

The @ap_clo@ method for function applications is also similarly mixing the two
previous versions.  We use the staged version of @lift_nd@ to lift the
next-stage set of functions into the monad stack.
For each closure @clo@ in the set, we generate a next-stage value, representing
the function application @clo(arg)@, by using @emit_ap_clo@.  Again, the method
@emit_ap_clo@ produces a current-stage representation of the future-stage
application result.  Finally, we reify the @out@ cache, store, and values,
which are all from the future-stage, back into the current-stage monadic value.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[ValSt], Cache)]
  def ap_clo(ev: Expr => Ans)(funs: Rep[Value], arg: Rep[Value]): Ans = for {
    σ   <- get_store;      clo <- lift_nd[AbsValue](funs)
    in  <- ask_in_cache;   out <- get_out_cache
    res <- lift_nd[(Set[ValSt], Cache)](Set(emit_ap_clo(clo, arg, σ, in, out)))
    _   <- put_out_cache(res._2);  vs <- lift_nd[ValSt](res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

@@

Particularly, if we are specializing a monovariant analysis, the address space
is equivalent to the set of all identifiers in the program. Utilizing this fact, the
result of accesses to the environment can be computed statically and we may
generate addresses directly during staging. After which, the store can be
specialized as an array by using the way mentioned above.

### 2019-09-09 paper/60_cases.tex
polish

In this section, we show AC can be understood and
implemented as an instance of staging abstract interpreters. We first revisit
the original implementation of abstract compilation for 0-CFA, and then
reproduce their result by simply adding stage annotations.
The program generated by our approach provides approximately the same
improvement with regards to speed, but without changing a single line of the
analyzer program using type-based staging annotations. However, AC requires
more engineering effort, i.e., rewriting the whole analyzer into the
closure-generation form. Moreover, as shown in Section~\ref{staged_ds}, our
approach is able to not only remove the interpretive overhead, but also enable
several optimizations such as specialized data structures.

@@

In this section, to demonstrate that the SAI approach is widely applicable, we
extend and refactor the staged abstract interpreter we presented in Section
\ref{sai} to various control-flow analyses (CFA). We first discuss the
calling-context sensitive analysis for staged analyzers by tweaking the
allocation strategy. Then, we describe how to implement store-widening with the
staged monads, which is a common optimization in CFA. Finally, we integrate
abstract garbage collection with staging. All of them are orthogonal to
staging.

@@

The abstract addresses we introduced in Section \ref{sai} are simply identical
to the variable names, which is a context-insensitive schema.  Through tweaking
the address allocation strategy, we can achieve various context-sensitive
analyses \cite{DBLP:conf/icfp/Gilray0M16}. In this section, we demonstrate a
calling-context sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, can be implemented in our staged abstract
interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

@@

A commonly-used technique to improve the running time of control-flow analyses
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, a store-widening analysis can be easily
implemented in the monadic abstract interpreter by swapping the @StateT@ and
the @SetT@ monad transformer. In this section, we show this optimization is
orthogonal to staging and we just need to similarly swap the same
\textit{staged} version of monads. After the adjustment, conceptually, we have
the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

The idea of abstract garbage collection is shown in Figure \ref{fixgc}. We
update the @fix@ function with caching (from Section \ref{sai}) to @fix_gc@
(the caching part is omitted in the presented code snippet). Several auxiliary
functions are used: @ask_root@ retrieves the current root addresses from the
monad stack; @root@ computes the reachable addresses given a value; and @gc@
performs garbage collection according to the given reachable addresses and
returns a new store. For the generic interpreter, we also need to compute the
root addresses for compound expressions; for the detail of this part, readers
may refer to \cite{DBLP:journals/pacmpl/DaraisLNH17}.

@@

One of the challenges of modern static analysis is that programs usually depend
on large libraries, and analyzing these libraries is time-consuming
\cite{toman_et_al:LIPIcs:2017:7121}.
If we are able to analyze programs and libraries separately and reuse the
results without losing precision, we can reduce the overhead of repeatedly
analyzing libraries. Some static analyzers compute a summary for a function or
a module, which can be reused later. However, when lacking the context
information, those analyses can be too conservative or unsound, which both lead
to imprecision. In this section, we show that meta-programming can provide a
trade-off between summary-based modular analysis and slow whole-program
analysis by utilizing an existing whole-program analyzer.

As we have already seen in Section \ref{sai}, the primitive
operation @close@ can denote a $\lambda$ term to a next-stage Scala function.
The specializing of the staged abstract interpreter to a $\lambda$ term is
independent to other parts of the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}

We use Scheme as an example, and consider a library that contains a
function @map@:
\begin{lstlisting}
  (define (map xs f) (if (null? xs) '() (cons (f (car xs)) (map (cdr xs) f))))
\end{lstlisting}

@@

We borrow the notation from traditional modular analysis
\cite{DBLP:conf/cc/CousotC02}: such a generated function is a modularly
reusable artifact, i.e., independent to its calling context.
The generated functions already eliminate the interpretation overhead and
monadic overhead, and can therefore be considered as partial summaries. However, we
still need to run the generated function at the next stage to obtain the analysis
result.  At the next stage, we can provide different abstract arguments, stores, and
caches, depending on the context where it is called.

>>>>

We show AC can be understood and implemented as an instance of staging abstract
interpreters. We first revisit the original implementation of abstract
compilation for 0-CFA, and then reproduce their result by simply adding stage
annotations.  The program generated by our approach provides approximately the
same improvement with regards to speed, but without changing a single line of
the analyzer program using type-based staging annotations. However, AC requires
more engineering effort, i.e., rewriting the whole analyzer into the
closure-generation form. Moreover, as shown in Section~\ref{staged_ds}, our
approach is able to not only remove the interpretive overhead, but also enable
several optimizations such as specialized data structures.

@@

To demonstrate that the SAI approach is applicable to various analyses, we now
extend and refactor the staged abstract interpreter we presented in Section
\ref{sai} to several variants of control-flow analyses (CFA). We first discuss
the calling-context sensitive analysis for staged analyzers by tweaking the
allocation strategy. Then, we describe how to implement store-widening with the
staged monads, which is a common optimization in CFA. Finally, we integrate
abstract garbage collection with staging. All of them are orthogonal to
staging.

@@

The abstract addresses we introduced in Section \ref{sai} are simply identical
to the variable names, which is a context-insensitive schema.  Through tweaking
the address allocation strategy, we can achieve various context-sensitive
analyses \cite{DBLP:conf/icfp/Gilray0M16}. Here, we demonstrate a
calling-context sensitive analysis, i.e., $k$-CFA-like analysis
\cite{DBLP:journals/jfp/HornM12}, can be implemented in our staged abstract
interpreter.
\begin{lstlisting}
  type Time = List[Expr]
  case class KCFAAddr(x: Ident, time: Time) extends Addr
\end{lstlisting}

@@

A commonly-used technique to improve the running time of control-flow analyses
is store-widening. As shown by \cite{Darais:2015:GTM:2814270.2814308,
DBLP:journals/pacmpl/DaraisLNH17}, a store-widening analysis can be easily
implemented in the monadic abstract interpreter by swapping the @StateT@ and
the @SetT@ monad transformer. This optimization is orthogonal to staging.
Within the monadic interpreter framework, we just need to similarly swap the
same \textit{staged} version of monads. After this adjustment, conceptually, we
have the following monad stack:
\begin{lstlisting}
  type AnsM[T] = RepReaderT[RepSetT[RepStateT[
                   RepReaderT[RepStateT[IdM, Cache, ?], Cache, ?], Store, ?], ?], Env, T]
\end{lstlisting}

@@

The skeleton of a staged version of abstract garbage collection implementation
is shown in Figure \ref{fixgc}.
We refactor the @fix@ function with caching (from Section \ref{sai}) to @fix_gc@
(the caching part is omitted in the presented code snippet). Several auxiliary
functions are used: @ask_root@ retrieves the current root addresses from the
monad stack; @root@ computes the reachable addresses given a value; and @gc@
performs garbage collection according to the given reachable addresses and
returns a new store. For the generic interpreter, we also need to compute the
root addresses for compound expressions; for the detail of this part, readers
may refer to \cite{DBLP:journals/pacmpl/DaraisLNH17}.

@@

One of the challenges of modern static analysis is that programs usually depend
on large libraries, and analyzing these libraries is time-consuming
\cite{toman_et_al:LIPIcs:2017:7121}.
If we are able to analyze programs and libraries separately and reuse the
results without losing precision, we can reduce the overhead of repeatedly
analyzing libraries. Some static analyzers compute a summary for a function or
a module, which can be reused later. However, when lacking the context
information, those analyses can be too conservative or unsound, which both lead
to imprecision. In this section, we show that meta-programming can provide a
trade-off between summary-based modular analyses and slow whole-program
analyses, by utilizing an existing whole-program analyzer.

As we have already seen in Section \ref{sai}, the primitive operation @close@
can denote a $\lambda$ term to a next-stage Scala function. Similar to
compilers, the specializing of the staged abstract interpreter to a $\lambda$
term is also independent of other parts of the analyzed program.
\begin{lstlisting}
  def close(ev: Eval => Ans)(λ: Expr, ρ: Rep[Env]): Rep[Value]
\end{lstlisting}

To illustrate the idea concretely, we consider a library that contains a
function @map@ and use the Scheme language as example:
\begin{lstlisting}
  (define (map xs f) (if (null? xs) '() (cons (f (car xs)) (map (cdr xs) f))))
\end{lstlisting}

@@

We borrow the notation from traditional modular analysis
\cite{DBLP:conf/cc/CousotC02}: such a generated function is a modularly
reusable artifact, i.e., independent of its context.
The generated functions already eliminate the interpretation overhead and
monadic overhead via staging, and can therefore be considered as partial
summaries. However, we still need to run the generated function at the next
stage to obtain the analysis result.  At the next stage, we can provide
different abstract arguments, stores, and caches, depending on the context
where it is called.

### 2019-09-09 paper/70_eval.tex
polish

To evaluate how performance can be improved by our staged abstract interpreters
approach, we scale our toy interpreter to support a large subset of the Scheme
language and evaluate the analysis time against the unstaged versions. The
result shows that the staged versions are, on average, 10+ times as fast as the
unstaged version, as evidence of our \textit{abstraction without regret}
approach.

>>>>

To evaluate how much performance can be improved by the staging, we scale our
toy staged abstract interpreter to support a large subset of the Scheme language and evaluate
the analysis time against the unstaged versions. The result shows that the
staged versions are, on average, 10+ times as fast as the unstaged version, as
evidence of our \textit{abstraction without regret} approach.
